Predicting complex diseases: performance and robustness [OF NEW METHOD?]

### 2018-03-02 paper/draft.tex
abstract + discussions + typos

\newpage
\abstract{
\textbf{Motivation:} \\
\textbf{Results:} \\
\textbf{Availability:} \\
\textbf{Contact:} \href{florian.prive@univ-grenoble-alpes.fr}{florian.prive@univ-grenoble-alpes.fr} \& \href{michael.blum@univ-grenoble-alpes.fr}{michael.blum@univ-grenoble-alpes.fr}\\
\textbf{Supplementary information:}
}

@@

Many methods have been developed in order to maximize the predictive power of genotypes for diseases, or more generally for phenotypes.
A commonly used technique, called ``P+T'' or ``C+T'' (which stands for ``Clumping and Thresholding'') -- or genetic profiling ..or even just PRS.. -- is used to derive PRSs from results of Genome-Wide Association Studies (GWASs) \cite[]{chatterjee2013projecting,dudbridge2013power,evans2009harnessing,purcell2009common,wray2007prediction}. This technique only use summary statistics which makes it very suitable and also very fast.
Linear Mixed-Models (LMMs) are another widely-used method in fields such as plant and animal breeding or for predicting highly heritable quantitative human phenotypes such as height \cite[]{lello2017accurate,yang2010common}. Yet, these models are not designed for predicting a binary trait such as a disease status and have proven to not be optimal at such task in another comparative study \cite[]{abraham2013performance}. Moreover, these methods and their derivatives are often computationally demanding, both in terms of memory and time requirements, which makes them unlikely to be used for prediction on large cohorts \cite[]{golan2014effective,maier2015joint,speed2014multiblup,zhou2013polygenic}.
Finally, statistical learning methods have also been used to derive PRSs for complex human disease by jointly estimating SNP effects. Such methods include logistic regression, Support Vector Machine (SVM) and random forests \cite[]{abraham2012sparsnp,abraham2014accurate,botta2013walk,botta2014exploiting,wei2009disease}.

@@

There have already been comparative studies about predicting human binary phenotypes based on genotype data. 
\citep{abraham2013performance} showed that penalized logistic regression and penalized SVM performed better than the standard C+T method and LMMs, for a wide-range of human diseases.
\cite{zhou2013polygenic} used simulations on real genotypes of Australian individuals to show that a hybrid of linear mixed models and sparse regression models performed better in a range of sparse to polygenic disease architectures. 
\cite{spiliopoulou2015genomic} investigated the effect of relatedness in prediction performance.
\cite{ware2017heterogeneity} evaluated best practices for using the C+T method and found out that the optimal inclusion threshold is trait-dependent. 
In this study, we particularly extend the work of \cite{abraham2013performance} by using a simulation framework similar but ..richer.. than the one used in \cite{zhou2013polygenic}. In order to make our comparison as comprehensive as possible, we compare different architectures of disease (number, size and location of causal effects and heritability) with different model for generating liability ..scores.., one with only linear effects, and one which combines linear, dominant and interaction effects. First, we quickly discard T-Trees as being the best method for predicting disease status based on genotypes.
Then, we show that penalized logistic regression consistently performs better than the C+T method whereas predictive performance of the C+T method is very sensitive to the threshold of inclusion of SNPs, depending of the architecture of disease, as shown in \cite{ware2017heterogeneity}.

@@

To simulate phenotypes, we use the Liability Threshold Model (LTM) with a prevalence of 30\% [CITATION?]. We vary different parameters: the number of chromosomes used (either all 22 chromosomes or only the chromosome 6), the size of the training set (..from 1000 to 6000 by 1000.., and the corresponding test set is composed of the remaining individuals), the number of causal variants and their location (30, 300 or 3000 anywhere on the dataset or 30 only in the HLA region), the heritability (50\% or 80\%), the distribution of effects associated with causal SNPs (Normal or Laplace), and the model used to generate the genetic ..part.. of the phenotypes (one ``simple model'' with only linear effects, and one ``fancy model'' which combines linear, dominant and interaction effects).
For the ``simple model'', we compute $$y_i = \sum_{j\in E_\text{causal}} w_j \cdot \widetilde{G_{i,j}}$$ where $w_j$ are weights (generated with a Gaussian or a Laplace distribution), $G_{i,j}$ is the allele count of individual $i$ for SNP $j$ and $\widetilde{G_{i,j}}$ corresponds to its standardized version (zero mean and unit variance for all SNPs). 
For the ``fancy model'', we separate the causal SNPs in three equal sets $E_\text{causal}^{(1)}$, $E_\text{causal}^{(2)}$ 
and $E_\text{causal}^{(3)}$ ($E_\text{causal}^{(3)}$ is further separated in two equal sets, $E_\text{causal}^{(3.1)}$ and $E_\text{causal}^{(3.2)}$). We then compute $$y_i = \underbrace{\sum_{j\in E_\text{causal}^{(1)}} w_j \cdot \widetilde{G_{i,j}}}_\text{linear} + \underbrace{\sum_{j\in E_\text{causal}^{(2)}} w_j \cdot \widetilde{D_{i,j}}}_\text{dominant} + \underbrace{\sum_{\substack{k=1 \\ j_1=e_k^{(3.1)} \\ j_2=e_k^{(3.2)}}}^{k=\left|E_\text{causal}^{(3.1)}\right|} w_{j_1} \cdot \widetilde{G_{i,j_1} G_{i,j_2}}}_\text{interaction}$$ where $D_{i,j} = \mathds{1}\left\{G_{i,j} \neq 0\right\}$ and $E_\text{causal}^{(q)}=\left\{e_k^{(q)},~k \in \left\{1, \ldots, \left|E_\text{causal}^{(q)}\right|\right\}\right\}$. Note that for the interaction part of the model, we scale interactions, not the raw allele counts, so that corresponding SNPs still display a marginal effect. 
For both models, the $y_i$ are then standardized to make sure they have a variance exactly equal to the desired heritability $h^2$ and we further add some environmental noise $\epsilon_i$ to $y_i$ where $\epsilon \sim N(0, 1 - h^2)$.

We make 3 different simulation scenarios, summarized in table \ref{tab:simus}. 
Scenario \textnumero1 uses the whole dataset (the 22 chromosomes) and a training set of size 6000. It compares all methods (see section \ref{sec:methods}) on all other combination of parameters in 100 simulations (i.e.\ 5 times more than in \cite{zhou2013polygenic}). Note that the T-Trees method is compared on only 5 simulations and an heritability of 80\%, which aims at quickly discarding this method that can require more that one hour of computation for one simulation. 
Scenario \textnumero2 consists in 100 simulations comparing the ``logit-simple'' and the C+T methods on a dataset composed of only the chromosome 6. Reducing the number of SNPs aims at increasing the ..polygenicity.. of our simulated models and at ..virtually.. increasing the sample size [REF? DUDBRIDGE?]. For this scenario, we use only the additive (``simple'') model, but continue to compare all previous variations of the other parameters.
Finally, scenario \textnumero3 reuses the whole dataset but varying the size of the training set in order to assess how the number of samples affects predictive performance of the PRS and the ``logit-simple'' methods. 100 simulations are run using 300 causal SNPs anywhere on the genome.

@@

In this study, we use two different measures of predictive accuracy. 
First, we use the Area Under the Receiver Operating Characteristic (ROC) Curve (AUC) \cite[]{fawcett2006introduction,hanley1982meaning,lusted1971signal}. In the case of our study, the AUC is the probability that a PRS of a case is greater than the PRS of a control.
This measure indicates how well we can distinguish between cases and controls using PRSs.
As a second measure, we report the percentage of the partial AUC for a specificity between 90\% and 100\% \cite[]{dodd2003partial,mcclish1989analyzing}. This measure is similar to the AUC, but focuses on high specificities (i.e.\ discarding most of the healthy population as being diseased) which is often the most useful part of the ROC curve in clinical settings.

@@

The C+T (Clumping + Thresholding) method directly derives a Polygenic Risk Score (PRS) from the results of Genome-Wide Associations Studies (GWASs), named summary statistics.
For the C+T method, a coefficient of regression is learned independently for each SNP along with a corresponding p-value (the GWAS part). 
The SNPs are first clumped (C) so that there remain only loci that are weakly correlated with each other. 
Then, thresholding (T) consists in removing SNPs that are within a certain level of significance (p-value threshold to be determined). 
Finally, a PRS is defined as the sum of allele counts of the remaining SNPs weighted by the corresponding effect coefficients: $$S_i(T, E_\text{clumping}) = \sum_{j \in E_\text{clumping}} \mathds{1}\{p_j < T\} \cdot \beta_j \cdot G_{i,j}$$ where $\beta_j$ ($p_j$) are the effect sizes (p-values) learned from the GWAS and $T$ is a threshold on p-values to be determined. In this study, we report scores for a clumping threshold at $r^2 > 0.2$ within regions of 500kb.

@@

T-Trees (\textit{Trees inside Trees}) is an algorithm derived from random forests \cite[]{breiman2001random} that takes into account the correlation structure among the genetic markers implied by linkage disequilibrium in GWAS data \cite[]{botta2014exploiting}. Since this method is significantly different from the other tested here and achieved surprisingly ..good.. results in the past, we decided to include T-Trees in this comparison in order to assess whether or not it could give superior predictive performance. We use the same parameters as reported in Table 4 of \cite{botta2014exploiting}. Yet, we use only 100 trees (instead of 1000) because it showed only a very subtle increase of AUC while ..taking way too much time.. (e.g.\ AUC of 81.5\% instead of 81\%, data not shown). %We call this method ``T-Trees'' in the results.

The last method we compare is penalized logistic regression.
We solve: $$\argmin_{\beta_0, \beta}(x, y, \lambda)\left\{\underbrace{\frac{1}{n}\sum_{i=1}^n \log\left(1+e^{-y_i (\beta_0+x_i^T\beta)}\right)}_\text{Loss function} + \underbrace{\lambda \left((1-\alpha)\frac{1}{2}\|\beta\|_2^2 + \alpha \|\beta\|_1\right)}_\text{Penalization}\right\}$$ where, in this study, $x$ is denoting the genotypes and covariables (e.g.\ principal components), $y$ is the disease status we want to predict and $\lambda$ is a regularization parameter that need to be determined. 
Different regularizations can be used to prevent overfitting, among other benefits: the L2-regularization (``ridge'', \cite{hoerl1970ridge}) shrinks coefficients and is ideal if there are many predictors drawn from a Gaussian distribution (corresponds to $\alpha = 0$ in the previous equation);
the L1-regularization (``lasso'', \cite{tibshirani1996regression}) forces some of the coefficients to be exactly zero and can be used as a means of variable selection, leading to sparse (and therefore more interpretable) models  (corresponds to $\alpha = 1$); 
the L1- and L2-regularization (``elastic-net'', \cite{zou2005regularization}) is a compromise between the two previous penalties and is particularly useful in the $m \gg n$ situation ($m$: number of SNPs), or any situation where there are many correlated predictors (corresponds to $0 < \alpha < 1$) \cite[]{friedman2010regularization}. In this study, we always use the elastic-net regularization with $\alpha = 0.5$, without trying to tune (e.g.\ by grid-search and cross-validation) the value of this hyper-parameter $\alpha$.

To fit this penalized logistic regression, we use a very efficient algorithm \cite[]{friedman2010regularization,tibshirani2012strong,zeng2017efficient} from which we derived our own implementation in our package bigstatsr \cite[]{prive2017efficient}. 
This type of algorithm builds predictions for many values of $\lambda$ (typically a ``regularization path'' of 100 values). To get an algorithm free of the choice of this hyper-parameter $\lambda$, we developed a procedure that we called Cross-Model Selection and Averaging (CMSA). First, this procedure separates the training set in $K$ folds (e.g.\ 10 folds). Secondly, in turn, each fold is considered as an inner validation set and the others ($K - 1$) folds form an inner training set, the model is trained on the inner training set and the corresponding predictions (scores) for the inner validation set are computed, the vector of scores which maximizes the likelihood on the inner validation set is determined and the vector of coefficients corresponding to the previous vector of scores is chosen. Finally, the $K$ resulting vectors of coefficients are reported and can be combined into one vector (e.g.\ using the geometric median) or resulting scores can be combined (e.g.\ averaging the $K$ probabilities for each individual). We also use this procedure to derive an early stopping criterion so that the algorithm does not need to evaluate the whole regularization path, making the whole procedure much faster.
Because of L1-regularization, this vector of coefficients is typically very sparse and can be used to make a PRS based on a \textit{linear} combination of allele counts. We call this method ``logit-simple'' in the results.

@@

In results, we use estimates of maximum achievable AUC values for the simulations: 84\% (respectively 94\%) for heritabilities of 50\% (respectively 80\%) based on three estimations (see Supplementary Materials). Note that we do not report the simulation results in terms of partial AUC because partial AUC values have a Spearman correlation of 98\% with the AUC results for all methods (Figure \ref{fig:supp-AUC-corr}).

@@

In scenario \textnumero1, we compared our penalized logistic regression (``logit-simple'') with the C+T method (``PRS'') on the whole genome (Table \ref{tab:simus}). 
Using the penalized logistic regression, we report AUCs greater than 93\%, nearly reaching the maximum achievable AUC of 94\%, which is not obtained by the upper bound of the C+T method in any scenario (Figures \ref{fig:main-AUC-logit} and \ref{fig:supp-AUC-logit}). 
Our penalized regression methods consistently provides better predictive performance than the C+T method for all disease architectures tested, excepted sometimes when we simulate 3000 causals SNPs anywhere on the genome, a case where all methods rarely exceeds 65\% of AUC (Figures \ref{fig:main-AUC-logit} and \ref{fig:supp-AUC-logit}). 
In the case where there are correlation between predictors (i.e.\ when we simulate 30 SNPs in the HLA region), the ``logit-simple'' method particularly performs better than the ``PRS-max'' one, with a mean AUC of 92.5\% versus 83\%, for a maximum achievable AUC of 94\% (Figure \ref{fig:main-AUC-logit}).
We also report that the choice of the threshold of inclusion of SNPs in the C+T method can greatly impact its performance, for example dropping from an AUC of 90\% to less than 60\% when including all SNPs passing clumping in a model with only 30 causal SNPs, or dropping from 60\% to a random predictor (50\% of AUC) when including only GWAS-significant SNPs in a model with 3000 causal SNPs (Figures \ref{fig:main-AUC-PRS} and \ref{fig:supp-AUC-PRS}).

@@

First, in scenario \textnumero1 we also tested the T-Trees method. 
For example, compared to our penalized logistic regression, T-Trees perform worse in term predictive measures, while taking much longer to run and making more complex predictive models because it uses more SNPs in the models and has non-linear effects (Figure \ref{fig:supp-ttrees}). Even when simulating a fancy model in which there is dominant and interaction effects that decision trees based methods should handle, AUC is still worse with T-Trees as compared with the ``logit-simple'' method.

Secondly, as for the comparison between our two penalized logistic regression in scenario \textnumero1, the classic one (called ``simple'') and the one using additional features coding for recessive and dominant effects (called ``triple''), we report that the ``triple'' method is nearly as good as the ``simple'' one when there are only linear effects and can lead to significantly greater results when there are also dominant and interactions effects (Figures \ref{fig:supp-triple} and \ref{fig:supp-AUC-triple}). Yet, the ``triple'' solution takes 2-3 more times to run and 3 more disk storage.

Thirdly, for the ``fancy'' simulation model, we report the exact same results as in the previous subsection that is to say that the ``logit-simple'' give greater AUCs than the ``PRS-max'', excepted in the case of 3000 causal SNP anywhere on the genome, where no method exceed 60\% of AUC (Figure \ref{fig:supp-AUC-logit-fancy}). Moreover, the choice of the threshold inclusion of SNPs also greatly affect the performance of the C+T method (Figure \ref{fig:supp-AUC-PRS-fancy}).

@@

These simulations corresponds to scenarios \textnumero2 and \textnumero3 (Table \ref{tab:simus}). First, when reproducing simulations of scenario \textnumero1 on only the chromosome 6, methods globally increase or ..stagnate.. in predictive performance (Figure \ref{fig:supp-AUC-chr6}). There is particularly a high increase in the case when simulating 3000 causal SNPs anywhere on the dataset: for example, AUC from the ``logit-simple'' can increase from less than 60\% to nearly 80\%. On the contrary, when simulating only 30 causal SNPs on the corresponding dataset, AUC of the ``PRS-max'' seems to slightly decrease. Secondly, when varying the training size, we report an increase of AUC when increasing the training size, with the AUC of the ``logit-simple'' method increasing faster than the ``PRS-max'' method (Figure \ref{fig:main-AUC-ntrain}).

@@

In this comparative study, we propose other methods as a replacement to the widely used C+T method. 
We show that jointly estimating the SNP effects using efficient penalized logistic regressions enable to achieve really high predictive performance, as reported in a previous study \cite[]{abraham2013performance}, which is not the case with the C+T method that uses SNP effects learned independently.
We are well aware that the C+T method is widely used because it is simple and can use summary statistics already learned from large genotype data, which is not the case for our penalized logistic regressions yet. 
However, in disease prediction for clinical interventions, best possible methods should be used, even if it has to be used by large consortia only [EVEN LARGE CONSORTIA DON'T HAVE ALL DATA?]. The logistic regression we present here could be used on data as large as biobank datasets because it is very efficient in both memory and computation time (..even faster than doing a GWAS!..).

%One does not simply identify high-risk individuals (Figure \ref{fig:meme}).

As for the C+T method itself, we recall that we report the maximum AUCs of 102 thresholds, a threshold that should normally be learned on the training set only. The choice of this threshold is very important as it can greatly impact the predictive performance of the method. In the penalized logistic regression we present, we remove the choice of the regularization parameter with a procedure that we call Cross-Model Selection and Averaging (CMSA) while keeping the same predictive performance.

>>>>

\newpage
[ABSTRACT TROP LONG?]
\abstract{
Polygenic Risk Scores (PRSs) consist in combining the information across many single-nucleotide polymorphisms (SNPs) in a score reflecting the genetic risk of developing a disease. PRSs can have a major public health impact, allowing for screening campaigns to identify high-genetic risk individuals for a given disease. Yet, current PRSs are far from reaching this goal, and would have to show a high discriminative power between cases and controls in order to be of any use.

The predictive power of PRS strongly relies on the underlying technique used to include SNPs in the PRS and to estimate their relative contribution. The most commonly used method to derive PRSs is called ``Clumping+Thresholding'' (C+T). It uses only univariate genome-wide association studies (GWAS) summary statistics, which makes it very suitable and also very fast. We recently developed two R packages, bigstatsr and bigsnpr, for the analysis of large-scale genome-wide data. Package bigstatsr includes efficient functions for computing penalized linear and logistic regressions on large datasets.

We present a comprehensive comparative study of the C+T method, our penalized logistic regression and the T-Trees algorithm, which is a derivation of random forests and has shown exceptionally good predictive results in a previous study. We quickly discard the T-Trees method and show that the penalized logistic regression consistently achieves higher predictive performance while being as fast as the C+T method, whereas predictive performance of the C+T method are very sensitive to the threshold of inclusion of SNPs, depending on the architecture of disease, as shown in previous studies.\\
\textbf{Contact:} \href{florian.prive@univ-grenoble-alpes.fr}{florian.prive@univ-grenoble-alpes.fr} \& \href{michael.blum@univ-grenoble-alpes.fr}{michael.blum@univ-grenoble-alpes.fr}\\
\textbf{Supplementary information:}
}

@@

Many methods have been developed to predict disease status, or more generally any phenotype, based on SNP information.
A commonly used method, called ``P+T'' or ``C+T'' (which stands for ``Clumping and Thresholding'') -- also sometimes called genetic profiling ..or even just PRS.. -- is used to derive PRSs from results of Genome-Wide Association Studies (GWASs) \cite[]{chatterjee2013projecting,dudbridge2013power,evans2009harnessing,purcell2009common,wray2007prediction}. This technique only use summary statistics which makes it very suitable and also very fast.
Linear Mixed-Models (LMMs) are another widely-used method in fields such as plant and animal breeding or for predicting highly heritable quantitative human phenotypes such as height \cite[]{lello2017accurate,yang2010common}. Yet, these models are not designed for predicting a binary trait such as a disease status and have proven to not be optimal at such task in another comparative study \cite[]{abraham2013performance}. Moreover, these methods and their derivatives are often computationally demanding, both in terms of memory and time requirements, which makes them unlikely to be used for prediction on large cohorts \cite[]{golan2014effective,maier2015joint,speed2014multiblup,zhou2013polygenic}.
Finally, statistical learning methods have also been used to derive PRSs for complex human disease by jointly estimating SNP effects. Such methods include logistic regression, Support Vector Machine (SVM) and random forests \cite[]{abraham2012sparsnp,abraham2014accurate,botta2013walk,botta2014exploiting,wei2009disease}.

@@

There have already been comparative studies about predicting human binary phenotypes based on genotype data. 
\citep{abraham2013performance} showed that penalized logistic regression and penalized SVM performed better than the standard C+T method and LMMs, for a wide-range of human diseases.
\cite{zhou2013polygenic} used simulations on real genotypes of Australian individuals to show that a hybrid of linear mixed models and sparse regression models performed better in a range of sparse to polygenic disease architectures. 
\cite{spiliopoulou2015genomic} investigated the effect of relatedness in prediction performance.
\cite{ware2017heterogeneity} evaluated best practices for using the C+T method and found out that the optimal inclusion threshold is trait-dependent. 
In this study, we particularly extend the work of \cite{abraham2013performance} by using a simulation framework similar but ..richer.. than the one used in \cite{zhou2013polygenic}. In order to make our comparison as comprehensive as possible, we compare different architectures of disease (number, size and location of causal effects and heritability) with different model for generating liability ..scores.., one with only linear effects, and one which combines linear, dominant and interaction effects. First, we quickly discard T-Trees as being the best method for predicting disease status based on genotypes.
Then, we show that the penalized logistic regression consistently achieves higher predictive performance while being as fast as the C+T method, whereas predictive performance of the C+T method are very sensitive to the threshold of inclusion of SNPs, depending on the architecture of disease, as shown in \cite{ware2017heterogeneity}.

@@

To simulate phenotypes, we use the Liability Threshold Model (LTM) with a prevalence of 30\% [CITATION?]. We vary different parameters: the number of chromosomes used (either all 22 chromosomes or only the chromosome 6), the size of the training set (..from 1000 to 6000 by 1000..), the number of causal variants and their location (30, 300 or 3000 anywhere on the genome or 30 only in the HLA region), the heritability (50\% or 80\%), the distribution of effects associated with causal SNPs (Normal or Laplace), and the model used to generate the genetic ..part.. of the phenotypes (one ``simple model'' with only linear effects, and one ``fancy model'' which combines linear, dominant and interaction effects).
For the ``simple model'', we compute $$y_i = \sum_{j\in E_\text{causal}} w_j \cdot \widetilde{G_{i,j}}$$ where $w_j$ are weights (generated with a Gaussian or a Laplace distribution), $G_{i,j}$ is the allele count of individual $i$ for SNP $j$ and $\widetilde{G_{i,j}}$ corresponds to its standardized version (zero mean and unit variance for all SNPs). 
For the ``fancy model'', we separate the causal SNPs in three equal sets $E_\text{causal}^{(1)}$, $E_\text{causal}^{(2)}$ 
and $E_\text{causal}^{(3)}$ ($E_\text{causal}^{(3)}$ is further separated in two equal sets, $E_\text{causal}^{(3.1)}$ and $E_\text{causal}^{(3.2)}$). We then compute $$y_i = \underbrace{\sum_{j\in E_\text{causal}^{(1)}} w_j \cdot \widetilde{G_{i,j}}}_\text{linear} + \underbrace{\sum_{j\in E_\text{causal}^{(2)}} w_j \cdot \widetilde{D_{i,j}}}_\text{dominant} + \underbrace{\sum_{\substack{k=1 \\ j_1=e_k^{(3.1)} \\ j_2=e_k^{(3.2)}}}^{k=\left|E_\text{causal}^{(3.1)}\right|} w_{j_1} \cdot \widetilde{G_{i,j_1} G_{i,j_2}}}_\text{interaction}$$ where $D_{i,j} = \mathds{1}\left\{G_{i,j} \neq 0\right\}$ and $E_\text{causal}^{(q)}=\left\{e_k^{(q)},~k \in \left\{1, \ldots, \left|E_\text{causal}^{(q)}\right|\right\}\right\}$. Note that for the interaction part of the model, we scale interactions, not the raw allele counts, so that corresponding SNPs still ..display.. a marginal effect. 
For both models, the $y_i$ are then standardized to make sure they have a variance exactly equal to the desired heritability $h^2$ and we further add some environmental noise $\epsilon_i$ to $y_i$, where $\epsilon \sim N(0, 1 - h^2)$.

We make 3 different simulation scenarios, summarized in table \ref{tab:simus}. 
Scenario \textnumero1 uses the whole dataset (the 22 chromosomes) and a training set of size 6000. It compares all methods (see section \ref{sec:methods}) on all other combination of parameters in 100 simulations (i.e.\ 5 times more than in \cite{zhou2013polygenic}). Note that the T-Trees method is compared on only 5 simulations and an heritability of 80\%, which aims at quickly discarding this method that can require more than one hour of computation for only one simulation. 
Scenario \textnumero2 consists in 100 simulations comparing the ``logit-simple'' and the C+T methods on a dataset composed of only the chromosome 6. Reducing the number of SNPs aims at increasing the ..polygenicity.. of our simulated models and at ..virtually.. increasing the sample size [REF? DUDBRIDGE?]. For this scenario, we use only the additive (``simple'') model, but continue to compare all previous variations of the other parameters.
Finally, scenario \textnumero3 reuses the whole dataset but varying the size of the training set in order to assess how the sample size affects predictive performance of the PRS and the ``logit-simple'' methods. 100 simulations are run using 300 causal SNPs anywhere on the genome.

@@

In this study, we use two different measures of predictive accuracy. 
First, we use the Area Under the Receiver Operating Characteristic (ROC) Curve (AUC) \cite[]{fawcett2006introduction,hanley1982meaning,lusted1971signal}. In the case of our study, the AUC is the probability that a PRS of a case is greater than the PRS of a control.
This measure indicates how well we can distinguish between cases and controls using PRSs.
As a second measure, the partial AUC for specificities between 90\% and 100\% is also reported \cite[]{dodd2003partial,mcclish1989analyzing}. This measure is similar to the AUC, but focuses on high specificities (i.e.\ discarding most of the healthy population as being diseased), which is often the most useful part of the ROC curve in clinical settings.

@@

The C+T (Clumping + Thresholding) method directly derives a Polygenic Risk Score (PRS) from the results of a Genome-Wide Associations Studies (GWAS), named summary statistics.
For the C+T method, a coefficient of regression is learned independently for each SNP along with a corresponding p-value (the GWAS part). 
The SNPs are first clumped (C) so that there remain only loci that are weakly correlated with each other. 
Then, thresholding (T) consists in removing SNPs that are within a certain level of significance (p-value threshold to be determined). 
Finally, a PRS is defined as the sum of allele counts of the remaining SNPs weighted by the corresponding effect coefficients: $$S_i(T, E_\text{clumping}) = \sum_{j \in E_\text{clumping}} \mathds{1}\{p_j < T\} \cdot \beta_j \cdot G_{i,j}$$ where $\beta_j$ ($p_j$) are the effect sizes (p-values) learned from the GWAS and $T$ is a threshold on p-values to be determined. In this study, we report scores for a clumping threshold at $r^2 > 0.2$ within regions of 500kb.

@@

T-Trees (\textit{Trees inside Trees}) is an algorithm derived from random forests \cite[]{breiman2001random} that takes into account the correlation structure among the genetic markers implied by linkage disequilibrium in GWAS data \cite[]{botta2014exploiting}. Since this method is significantly different from the other presented methods and achieved surprisingly high predictive performance in the past, we decided to include T-Trees in this comparison in order to assess whether or not it could give superior predictive performance. We use the same parameters as reported in Table 4 of \cite{botta2014exploiting}. Yet, we use only 100 trees (instead of 1000) because it showed only a very subtle increase of AUC while ..taking way too much time.. (e.g.\ AUC of 81.5\% instead of 81\%, data not shown). %We call this method ``T-Trees'' in the results.

The last method we compare is a penalized logistic regression.
We solve: $$\argmin_{\beta_0, \beta}(x, y, \lambda)\left\{\underbrace{\frac{1}{n}\sum_{i=1}^n \log\left(1+e^{-y_i (\beta_0+x_i^T\beta)}\right)}_\text{Loss function} + \underbrace{\lambda \left((1-\alpha)\frac{1}{2}\|\beta\|_2^2 + \alpha \|\beta\|_1\right)}_\text{Penalization}\right\}$$ where, in this study, $x$ is denoting the genotypes and covariables (e.g.\ principal components), $y$ is the disease status we want to predict and $\lambda$ is a regularization parameter that need to be determined. 
Different regularizations can be used to prevent overfitting, among other benefits: the L2-regularization (``ridge'', \cite{hoerl1970ridge}) shrinks coefficients and is ideal if there are many predictors drawn from a Gaussian distribution (corresponds to $\alpha = 0$ in the previous equation);
the L1-regularization (``lasso'', \cite{tibshirani1996regression}) forces some of the coefficients to be exactly zero and can be used as a means of variable selection, leading to sparse (and therefore more interpretable) models  (corresponds to $\alpha = 1$); 
the L1- and L2-regularization (``elastic-net'', \cite{zou2005regularization}) is a compromise between the two previous penalties and is particularly useful in the $m \gg n$ situation ($m$: number of SNPs), or any situation where there are many correlated predictors (corresponds to $0 < \alpha < 1$) \cite[]{friedman2010regularization}. In this study, we always use the elastic-net regularization with $\alpha = 0.5$, without trying to tune (e.g.\ by grid-search and cross-validation) the value of this hyper-parameter $\alpha$.

To fit this penalized logistic regression, we use a very efficient algorithm \cite[]{friedman2010regularization,tibshirani2012strong,zeng2017efficient} from which we derived our own implementation in the R package bigstatsr \cite[]{prive2017efficient}. 
This type of algorithm builds predictions for many values of $\lambda$ (typically a ``regularization path'' of 100 values). To get an algorithm free of the choice of this hyper-parameter $\lambda$, we developed a procedure that we called Cross-Model Selection and Averaging (CMSA). First, this procedure separates the training set in $K$ folds (e.g.\ 10 folds). Secondly, in turn, each fold is considered as an inner validation set and the others ($K - 1$) folds form an inner training set, the model is trained on the inner training set and the corresponding predictions (scores) for the inner validation set are computed, the vector of scores which maximizes the likelihood on the inner validation set is determined and the vector of coefficients corresponding to the previous vector of scores is chosen. Finally, the $K$ resulting vectors of coefficients are reported and can be combined into one vector (e.g.\ by using the geometric median) or resulting scores can be combined (e.g.\ by averaging the $K$ probabilities for each individual). 
We also use this procedure to derive an early stopping criterion so that the algorithm does not need to evaluate the whole regularization paths, making the whole procedure much faster.
Because of L1-regularization, this vector of coefficients is typically very sparse and can be used to make a PRS based on a \textit{linear} combination of allele counts. We call this method ``logit-simple'' in the results.

@@

In ..the.. results, we use estimates of maximum achievable AUC values for the simulations: 84\% (94\%) for heritabilities of 50\% (80\%) based on three estimations (see Supplementary Materials). Note that we do not report the simulation results in terms of partial AUC because partial AUC values have a Spearman correlation of 98\% with the AUC results for all methods (Figure \ref{fig:supp-AUC-corr}).

@@

In scenario \textnumero1, we compared our penalized logistic regression (``logit-simple'') with the C+T method (``PRS'') on the whole genome (Table \ref{tab:simus}). 
Using the penalized logistic regression, we report AUCs greater than 93\%, nearly reaching the maximum achievable AUC of 94\%, which is not obtained by the upper bound of the C+T method in any scenario (Figures \ref{fig:main-AUC-logit} and \ref{fig:supp-AUC-logit}). 
The penalized logistic regression consistently provides better predictive performance than the C+T method for all disease architectures tested, excepted sometimes when we simulate 3000 causals SNPs anywhere on the genome, a case where all methods rarely exceeds 65\% of AUC (Figures \ref{fig:main-AUC-logit} and \ref{fig:supp-AUC-logit}). 
In the case where there are correlation between predictors (i.e.\ when we simulate 30 SNPs in the HLA region), the ``logit-simple'' method particularly performs better than the ``PRS-max'' one, with a mean AUC of 92.5\% versus 83\%, for a maximum achievable AUC of 94\% (Figure \ref{fig:main-AUC-logit}).
We also report that the choice of the threshold of inclusion of SNPs in the C+T method can greatly impact its performance, for example dropping from an AUC of 90\% to less than 60\% when including all SNPs passing clumping in a model with only 30 causal SNPs, or dropping from 60\% to a random predictor (AUC of 50\%) when including only GWAS-significant SNPs in a model with 3000 causal SNPs (Figures \ref{fig:main-AUC-PRS} and \ref{fig:supp-AUC-PRS}).

@@

First, in scenario \textnumero1, we also tested the T-Trees method. 
For example, as compared to the ``logit-simple'' method, T-Trees perform worse in terms of predictive ability, while taking much longer to run and making more complex predictive models because it uses more predictors and non-linear effects (Figure \ref{fig:supp-ttrees}). Even when simulating a fancy model in which there are dominant and interaction effects that decision trees based methods should handle, AUC is still lower with T-Trees as compared with the ``logit-simple'' method.

Secondly, as for the comparison between the two penalized logistic regressions in scenario \textnumero1, the classic one (called ``simple'') and the one using additional features coding for recessive and dominant effects (called ``triple''), we report that the ``triple'' method is nearly as good as the ``simple'' one when there are only linear effects and can lead to significantly greater results when there are also dominant and interactions effects (Figures \ref{fig:supp-triple} and \ref{fig:supp-AUC-triple}). Yet, the ``triple'' solution takes 2-3 more times to run and 3 more disk storage.

Thirdly, for the ``fancy'' simulation model, we report the exact same results as in the previous subsection that is to say that the ``logit-simple'' achieves greater AUCs than the ``PRS-max'', excepted in the case of 3000 causal SNP anywhere on the genome, where no method exceed 60\% of AUC (Figure \ref{fig:supp-AUC-logit-fancy}). Moreover, the choice of the threshold inclusion of SNPs also greatly ..affects/impacts.. the performance of the C+T method (Figure \ref{fig:supp-AUC-PRS-fancy}).

@@

These simulations correspond to scenarios \textnumero2 and \textnumero3 (Table \ref{tab:simus}). First, when reproducing simulations of scenario \textnumero1 on only the chromosome 6, methods globally increase or ..stagnate.. in predictive performance (Figure \ref{fig:supp-AUC-chr6}). There is particularly a high increase in the case when simulating 3000 causal SNPs anywhere on the dataset: for example, AUC from the ``logit-simple'' can increase from less than 60\% to nearly 80\%. On the contrary, when simulating only 30 causal SNPs on the corresponding dataset, AUC of the ``PRS-max'' seems to slightly decrease. Secondly, when varying the training size, we report an increase of AUC when increasing the training size, with the AUCs provided by the ``logit-simple'' method increasing faster than for the ``PRS-max'' method (Figure \ref{fig:main-AUC-ntrain}).

@@

In this comparative study, we propose other methods as a replacement to the widely used C+T method. 
We show that jointly estimating the SNP effects using efficient penalized logistic regressions enables to achieve really high predictive performance, as reported in a previous study \cite[]{abraham2013performance}, which is not the case with the C+T method that uses SNP effects learned independently.
We are well aware that the C+T method is widely used because it is simple and can use summary statistics already learned from large genotype data, which is not the case for our penalized logistic regressions yet. 
However, in order to accurately predict disease status for clinical interventions, best possible methods should be used. The logistic regression we present in this paper could be used on data as large as biobank datasets because it is very efficient in both memory and computation time (..even faster than doing a GWAS!..).

As for the C+T method itself, we recall that we report the maximum AUC of 102 thresholds, a threshold that should normally be learned on the training set only. The choice of this threshold is very important as it can greatly impact the predictive performance of the method. On the contrary, in the penalized logistic regression we present, we remove the choice of the regularization parameter with a procedure that we call Cross-Model Selection and Averaging (CMSA) while keeping high predictive performance.
